---
title: "Breanna_Code"
author: "Tom Bird"
date: "29/09/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preamble

We are going to use a markdown document to communicate draft code and results, to try and setup a template for how we would report on these data to a client. 


## Load packages
I try and load packages separately from data.

```{r load_packages}
library(dplyr)
library(lme4)

```

## Loading data

In future, we should try and work from either a formatted database or a .csv extracted from that formatted database, and those data should have metadata etc etc.  For now we go with what we have.

```{r load_data}
#read saved data
Metrics <- readRDS("Data/Metrics.rds")
```

```{r view_raw_data}
head(Metrics)
```


## Checking distributional assumptions
Here we would put put some text describing the results, and setting up the next section, in case there were transformations we wanted to do.

```{r plot_histograms}
par(mfrow=c(2,4))

hist(Metrics$`Total Abundance (ind/0.1m²)`)

hist(Metrics$`Total Richness`)

hist(log(Metrics$`Total biomass (g/0.1m²)`))

hist(Metrics$`Simpson's Diversity (1-D)`)

hist(Metrics$`Simpson's Evenness (E)`)

hist(Metrics$`Swartz Dominance Index (SDI)`)

hist(Metrics$`Percentage of Axinopsida with low shell staining`)
```
 



##Explore Variables
Here were are looking at what the variables are relative to our indicators, and whether we want to add variable transformations.

###Asses if trends are reasonably linear

```{r plot_stuff}
Metrics %>%

  select(DiffuserDistance, DiffuserDirection, `Total Abundance (ind/0.1m²)`) %>%

  plot()

Metrics %>%

  select(DiffuserDistance, DiffuserDirection, `Total Richness`) %>%

  plot()

Metrics %>%

  select(DiffuserDistance, DiffuserDirection, `Total biomass (g/0.1m²)`) %>%

  plot()

Metrics %>%

  select(DiffuserDistance, DiffuserDirection, `Simpson's Diversity (1-D)`) %>%

  plot()

Metrics %>%

  select(DiffuserDistance, DiffuserDirection, `Simpson's Evenness (E)`) %>%

  plot()

Metrics %>%

  select(DiffuserDistance, DiffuserDirection, `Swartz Dominance Index (SDI)`) %>%

  plot()

Metrics %>% 

  select(DiffuserDistance, DiffuserDirection, `Percentage of Axinopsida with low shell staining`) %>%

  plot()
```
 

## modeling section
```{r models}
# Example model
 
summary(glmer(`Total Abundance (ind/0.1m²`) ~ DiffuserDirection + DiffuserDistance + DiffuserDirection:DiffuserDistance + (1|Rep), data = Metrics, family = poisson)
```


