---
title: "Breanna_Code"
author: "Tom Bird"
date: "29/09/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preamble

Here we are going to use a markdown document to communicate draft code and results, to try and setup a template for how we would report on these data to a client.

## Load packages

I try and load packages separately from data.

```{r load_packages}
library(dplyr)
library(lme4)
library(tidyverse)
library(lattice)
```

## Loading data

In future, we should try and work from either a formatted database or a .csv extracted from that formatted database, and those data should have metadata etc etc. For now we go with what we have.

```{r load_data}
#read saved data
Metrics <- readRDS("Data/Metrics.rds")

axinopsida <- readRDS("Data/axinopsida.rds")

Metrics <- Metrics %>%
  merge(axinopsida, by = c("Station", "Rep"))
```

## View raw data

Note that the raw data contain characters that will make things tricky later on. I would exclude parentheses etc. as general practice. Also names are quite long, making it difficult to include in the models. This is where having a data dictionary is useful, as you can keep descriptions, formal names, plot names available for use in figures.

```{r view_raw_data}
head(Metrics)
```

```{r data_dictionary}
#Putting a data dictionary here in case it is useful
dict=data.frame(longName=names(Metrics), shortName=c("Station", "Rep", "Abundance", "Richness", "Biomass", "Diversity", "Evenness", "Dominance", "Shell_Staining", "Year", "Distance", "Direction", "Shell_Successes", "Shell_Trials"))

#new data file with short names for playing
metrics <- Metrics %>%
  mutate(DiffuserDirection=ifelse(DiffuserDirection==1, "north", "south"))
names(metrics)=dict$shortName
```

## Checking distributional assumptions

Here we would put put some text describing the results, and setting up the next section, in case there were transformations we wanted to do.

##Explore Variables: Here were are looking at what the variables are relative to our indicators, and whether we want to add variable transformations.

### Check distribution of response variables

Determine the link and variance function (glm family) that will be used to transform the response variable for a good model fit. Count data or right skewed data is often poisson distributed.

Quassi families, such as quassipoisson, are used when the variance is proportional to the mean and may be used when data are underdispersed or overdispersed.

```{r histograms-raw}

metrics %>% 
  select(!c('Distance')) %>%
  pivot_longer(!c( 'Rep', 'Station', 'Year', 'Direction'), names_to="Indicator", values_to="value") %>%
  ggplot(aes(x=value)) +
    geom_histogram(aes(fill=Direction), bins=15)+
    facet_wrap(~Indicator, scale="free_x")


```

Transformed distributions

```{r histograms-transformed}
#abundance, biomass, dominance, and shell staining approximately poisson distributed. Shell staining may be better represented with negative binomial


metrics <- metrics %>%
  mutate(
#log_Shell_Staining = log(Shell_Staining+1),
         log_Biomass = log(Biomass), 
         .keep = "unused")

metrics %>% 
  select(!c('Distance')) %>%
  pivot_longer(!c( 'Rep', 'Station', 'Year', 'Direction'), names_to="Indicator", values_to="value") %>%
  ggplot(aes(x=value)) +
    geom_histogram(aes(fill=Direction), bins=10)+
    facet_wrap(~Indicator, scale="free_x")

```

## Checking response trends

Asses if trends are reasonably linear. Here we should be plotting response variables against predictors.

### Distance

Definitely some trends with distance, but also definitely some non-linear things happening.

```{r scatterplots}
#Distance scatterplot

metrics %>% 
  #select(!c('Direction')) %>%
  pivot_longer(!c( 'Rep', 'Station', 'Year', 'Distance', 'Direction'), names_to="Indicator",  values_to="value") %>%
  ggplot(aes(x=Distance, y=value, colour=Direction)) +
    geom_point()+
    stat_smooth(method = "lm", formula = y~x, aes(colour=Direction)) +
    facet_wrap(~Indicator, scale="free_y")

```

### Direction

```{r Direction boxplot}

metrics %>% 
  select(!c('Distance')) %>%
  pivot_longer(!c( 'Rep', 'Station', 'Year', 'Direction'), names_to="Indicator", values_to="value") %>%
  ggplot(aes(x=Direction, y=value, fill=Direction)) +
    geom_boxplot()+
    facet_wrap(~Indicator, scale="free_y")
```

## Modeling section

Only use model selection criteria like the AIC to compare models of the same family.

Note that GLMMs "are surprisingly challenging to use even for statisticians. Although several software packages can handle GLMMs (Table 1), few ecologists and evolutionary biologists are aware of the range of options or of the possible pitfalls. In reviewing papers in ecology and evolution since 2005 found by Google Scholar, 311 out of 537 GLMM analyses (58%) used these tools inappropriately in some way"

```{r abundance}
# Example model
# Abundance

# Initial model with interaction term and random effects. Asess dispersion using possion family.
modelabundance1=glmer(Abundance ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics, family = poisson)
summary(modelabundance1)


#test for overdispersion 
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

overdisp_fun(modelabundance1)
# p-value is small, likely over dispersed data. try quasipoisson to deal with overdispersion 

# fit model using non-binomial family
modelabundance2=glmer.nb(Abundance ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics)
summary(modelabundance2)
#overdisp_fun(modelabundance2)
#dispersion is better using negative binomial. Conclude that we can continue with poisson (simpler) but note that standard errors could be biased.




# Model with no random effects
modelabundance3=glm(Abundance ~ Direction + Distance + Direction:Distance,  data = metrics, family = poisson)
summary(modelabundance3)

# Compare both models AIC
summary(modelabundance1)$AIC[1] # with random effects
summary(modelabundance3)$aic    # without random effects
# conclude that model with random effects is better due to smaller AIC


#Remove non-significant interaction term from random effects model
modelabundance4=glmer(Abundance ~ Direction + Distance + (1|Station), data = metrics, family = poisson)
summary(modelabundance4)

# Compare both random effects models (with and without interaction)
summary(modelabundance1)$AIC[1] #with interaction
summary(modelabundance4)$AIC[1] #without interaction
# The model without the interaction term has a slightly higher AIC, but since the difference between the two models is less than 2 AIC units, the model without the interaction term is likely the better model.



#assess model 4



# diagnostic plot
plot(modelabundance4,type=c("p","smooth"))
  # type p is the points of the scatterplot, 
  # smooth is the locally estimated scatterplot smoothing (loess)

  # note that heteroscedasticity looks ok for model residuals
  # note that residuals look mostly normally distributed in the qqplot, 
  # with some heavy tails. Should be ok as glm can deal with heavy tails?
  # smooth curve is mostly straight, looks like model fit is ok

# QQ normal plot
qqnorm(resid(modelabundance4), main="normal qq-plot, residuals")
qqline(resid(modelabundance4))

library(lattice)

# plot the random effects
dotplot(ranef(modelabundance4,condVar=TRUE)) 
  # ranef to extract estimates of random effects, 
  # dotplot quickly plots the estimates from the model
  # station 14 is the most extreme

```

```{r richness}
# Richness

# Initial model with interaction term and random effects. Asess dispersion using possion family.
modelrichness1=glmer(Richness ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics, family = poisson)
summary(modelrichness1)


#test for overdispersion 
overdisp_fun <- function(model) {
    rdf <- df.residual(model)
    rp <- residuals(model,type="pearson")
    Pearson.chisq <- sum(rp^2)
    prat <- Pearson.chisq/rdf
    pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
    c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}


overdisp_fun(modelrichness1)
# p-value is large, likely not overdispersed data.




# Model with no random effects
modelrichness2=glm(Richness ~ Direction + Distance + Direction:Distance,  data = metrics, family = poisson)
summary(modelrichness2)

# Compare both models AIC
summary(modelrichness1)$AIC[1] # with random effects
summary(modelrichness2)$aic    # without random effects
# conclude that model without random effects is better due to smaller AIC
# keep significant interaction term 
# model2 is best


#assess model
library(lattice)


# diagnostic plots
plot(modelrichness2)
  # note that heteroscedasticity looks ok for model residuals
  # note that residuals look mostly normally distributed, 
  # with some heavy tails but overall looks good
  # smooth curve is mostly straight, looks like model fit is ok, 
  # there is a bit of a w shaped pattern, however
```

```{r biomass}
# Biomass

# biomass is not count data, poisson glm should not be used
# instead, start with lmer on transformed data
# Initial model with interaction term and random effects

modelbiomass1=lmer(log_Biomass ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics)
anova(modelbiomass1)
summary(modelbiomass1)

# plot the random effects
dotplot(ranef(modelbiomass1,condVar=TRUE)) 
  # ranef to extract estimates of random effects, 
  # dotplot quickly plots the estimates from the model
  # station 14 and 15 are the most extreme but overlap at 0, assume that random effect is not needed, but will test for it below


# Model with no random effects
modelbiomass2=lm(log_Biomass ~ Direction + Distance + Direction:Distance,  data = metrics)
summary(modelbiomass2)

# Compare both models AIC
anova(modelbiomass1, modelbiomass2)
# Conclude that random effects are not significant in the model.
# Also, the simpler model has lower AIC. 
# Removing random effects also simplifies interpretation of model. 

# Interaction not significant.
# Distance not significant
# Direction is significant
# Overall, there is little evidence that distance
# from the diffuser effects the biomass, 
# but there is some evidence that direction from the diffuser has an effect 


#Remove non-significant interaction term from fixed effects model
modelbiomass3=lm(log_Biomass ~ Direction + Distance, data = metrics)
summary(modelbiomass3)

# Compare both random effects models (with and without interaction)
anova(modelbiomass3, modelbiomass2)
# No evidence that the coefficient of theinteraction term is significant. 
# adjusted R squared is higher in model 3
# conclude that model 3 is the best model, without random effects or interactions

# Conclude that Direction is significant, 
# Distance not significant in the model used to
# describe biomass changes in study are in relation to the position of the diffuser
# Direction is significant in the model to predict biomass in relation to the diffuser

# Diagnostic plots
plot(modelbiomass3)
# possibly some evidence of heteroscadiscity, but minor
# some evidence of heavy-tailed residulas in the qq plot, but minor





```

```{r dominance}
# Swartz Dominance Index
# Initial model with interaction term and random effects. Asess dispersion using possion family.
modeldominance1=glmer(Dominance ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics, family = poisson)
summary(modeldominance1)
# r shows a warning, random effects are small

# Model with no random effects
modeldominance2=glm(Dominance ~ Direction + Distance + Direction:Distance,  data = metrics, family = poisson)
summary(modeldominance2)

# Compare both models AIC
summary(modeldominance1)$AIC[1] # with random effects
summary(modeldominance2)$aic    # without random effects
# conclude that model without random effects is better due to smaller AIC, 
# and is the simpler model

# all terms are significant in the model to describe dominance 
# in relation to diffuser

#assess model

# diagnostic plots
plot(modeldominance2)
  # note that heteroscedasticity looks ok for model residuals
  # note that residuals look mostly normally distributed in the qqplot, 
  # with some heavy tails. Should be ok as glm can deal with heavy tails?
  # smooth curve is mostly straight, looks like model fit is ok


```

```{r diversity}

# Diversity (Simpson's)

# diversity is not count data, poisson glm should not be used
# instead, start with lmer
# Initial model with interaction term and random effects

modeldiversity1=lmer(Diversity ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics)
anova(modeldiversity1)
summary(modeldiversity1)

# plot the random effects
dotplot(ranef(modeldiversity1,condVar=TRUE)) 
  # ranef to extract estimates of random effects, 
  # dotplot quickly plots the estimates from the model
  # station 14 and 13 are the most extreme, most overlap 0, random effect may be needed, will test for it below


# Model with no random effects
modeldiversity2=lm(Diversity ~ Direction + Distance + Direction:Distance,  data = metrics)
summary(modeldiversity2)

# Compare both models AIC
anova(modeldiversity1, modeldiversity2)
# Conclude that random effects are significant in the model.
# However, the simpler model has lower AIC. 
# Removing random effects would simplify interpretation 

# For the model with random effects, Distance, as well as the interaction of distance and direction are non-zero.
#Therefore, Distance and direction are important in determining diversity in relation to the diffuser



# Diagnostic plots
plot(modeldiversity1, type=c("p","smooth"))
  # type p is the points of the scatterplot, 
  # smooth is the locally estimated scatterplot smoothing (loess)
  # note that heteroscedasticity looks ok for model residuals
  # smooth curve is mostly straight, looks like model fit is ok

# QQ normal plot
qqnorm(resid(modeldiversity1), main="normal qq-plot, residuals")
qqline(resid(modeldiversity1))

  # note that residuals look mostly normally distributed in the qqplot, 
  # with some heavy tails on the lower left.
 







```

```{r evenness}
# Evenness (Simpson's)

# evenness is not count data, poisson glm should not be used
# instead, start with lmer 
# Initial model with interaction term and random effects

modelevenness1=lmer(Evenness ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics)
anova(modelevenness1)
summary(modelevenness1)

# plot the random effects
dotplot(ranef(modelevenness1,condVar=TRUE)) 
  # ranef to extract estimates of random effects, 
  # dotplot quickly plots the estimates from the model
  # station 14 and 13 are the most extreme, random effect is likely moderately important for a few sites, but will test for it below


# Model with no random effects
modelevenness2=lm(Evenness ~ Direction + Distance + Direction:Distance,  data = metrics)
summary(modelevenness2)

# Compare both models AIC
anova(modelevenness1, modelevenness2)
# Conclude that random effects are significant in the model.
# However, the simpler model has lower AIC. 
# Removing random effects would simplify interpretation 

# For the model with random effects, Distance, as well as the interaction of distance and direction are non-zero.
#Therefore, Distance and direction are important in determining evenness in relation to the diffuser



# Diagnostic plots
plot(modelevenness1, type=c("p","smooth"))
  # type p is the points of the scatterplot, 
  # smooth is the locally estimated scatterplot smoothing (loess)
  # note that heteroscedasticity looks ok for model residuals
  # smooth curve is mostly straight, looks like model fit is ok

# QQ normal plot
qqnorm(resid(modelevenness1), main="normal qq-plot, residuals")
qqline(resid(modelevenness1))
  # note that residuals look mostly normally distributed in the qqplot, 

 

```

```{r axinopsida}

# Since staining data are a proportion, use a binomial family glm. Included original counts of data rather than proportions

# Shell_Staining

# Initial model with interaction term and random effects. 
modelaxinopsida1=glmer(cbind(Shell_Successes, Shell_Trials - Shell_Successes) ~ Direction + Distance + Direction:Distance + (1|Station), data = metrics, family = binomial)
summary(modelaxinopsida1)

# Model with no random effects
modelaxinopsida2=glm(cbind(Shell_Successes, Shell_Trials - Shell_Successes) ~ Direction + Distance + Direction:Distance,  data = metrics, family = binomial)
summary(modelaxinopsida2)

# Compare both models AIC
summary(modelaxinopsida1)$AIC[1] # with random effects
summary(modelaxinopsida2)$aic    # without random effects
# conclude that model with random effects is better due to smaller AIC


#Remove non-significant interaction term from random effects model
modelaxinopsida3=glmer(cbind(Shell_Successes, Shell_Trials - Shell_Successes)~ Direction + Distance + (1|Station), data = metrics, family = binomial)
summary(modelaxinopsida3)

# Compare both random effects models (with and without interaction)
summary(modelaxinopsida1)$AIC[1] #with interaction
summary(modelaxinopsida3)$AIC[1] #without interaction
# The model without the interaction term has lower higher AIC and is a simpler model. This model without the interaction term is likely the better model.



# diagnostic plot
plot(modelaxinopsida3,type=c("p","smooth"))
  # type p is the points of the scatterplot, 
  # smooth is the locally estimated scatterplot smoothing (loess)

  # note that heteroscedasticity looks ok for model residuals
  # note that residuals look mostly normally distributed in the qqplot, 
  # with some heavy tails. Should be ok as glm can deal with heavy tails?
  # smooth curve is mostly straight, looks like model fit is ok

# QQ normal plot
qqnorm(resid(modelaxinopsida3), main="normal qq-plot, residuals")
qqline(resid(modelaxinopsida3))


# plot the random effects
dotplot(ranef(modelaxinopsida3,condVar=TRUE)) 
  # ranef to extract estimates of random effects, 
  # dotplot quickly plots the estimates from the model
  # station 5 and 4 are the most extreme
  # Random effect is likely required for a few sites

```
